Split-Apply-Combine with Dynamic Grouping

### 2023-10-03 paper/article.Rnw
added spaces around *

cumulate(input, collapse = A*B ~ A*B1 + A
        , test = function(d) nrow(d) >= 3, muY = mean(Y) )
@
%
Consider the formula \code{A*B ~ A*B1 + A} in the call to \code{cumulate()}.
The left-hand-side \code{A*B} is the target output grouping. The
right-hand-side is to be interpreted as the collapsing sequence: if an instance
of \code{A*B} does not pass the test, then collapse to \code{A*B1}, and if that
does not pass the test collapse to \code{A}. If this final grouping also does
not pass the test, the result is \code{NA}.

@@

It is possible to get the same result with \code{accumulate()}. This will cause
summation over all variables that are not used in the formula object. In the
below example we also introduce the helper function \code{min_records()}.
<<>>=
input$Y2 <- 11:19
accumulate(input, collapse = A*B ~ A*B1 + A
          , test = min_records(3), fun = mean)
@
This means that for experimentation, users must be careful to exclude categorical
variables that are not used from the input data set. For example, if \code{B1}
is not used, we get the following.
<<>>=
accumulate(input[-3], collapse = A*B ~ A
          , test = min_records(3), fun = mean)
@

@@

We will apply these conditions to the \code{input} dataset that was 
constructed in Section~\ref{sect:formula}. As a reminder we print
the first few records.
<<>>=
head(input, 4)
@
We use \code{A*B ~ A*B1 + B1} as collapsing scheme. The function
\code{from_validator} passes the requirements as a test function to
\code{accumulate()} (or \code{cumulate()}).
<<>>=
accumulate(input, collapse = A*B ~ A*B1 + B1,
    test = from_validator(rules), fun = mean)
@
Note that for target groups $(A=3,B=21)$ and $(A=3,B=22)$ none of the available
collapsing levels lead to a group that satisfied all conditions. Therefore the
collapsing level and output variables are all missing (\code{NA}).

The third and most flexible way for users to express tests is to write a custom
testing function. The requirements are that it must work on any subset of a
data frame, including a dataset with zero rows. The previous example can
thus also be expressed as follows.
<<>>=
my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >= 3
accumulate(input, collapse = A*B ~ A*B1 + B1, test = my_test, fun = mean)
@

@@

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation that includes the target subpopulation
\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
by simple random sampling from the population, the mean can be estimated with
the sample mean. In this example we will demand that there are at least ten
records for which turnover has been measured. The collapsing scheme is given by
\code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
classification by respectively the first two SBI digits and the first SBI
digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi, 1, 2)
producers$sbi1 <- substr(producers$sbi, 1, 1)
head(producers, 3)
@
Using \code{cumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers
             , collapse = sbi*size ~ sbi + sbi2 + sbi1
             , test = min_complete(n = 10, vars = "industrial")
             , mean_industrial = mean(industrial, na.rm = TRUE))
head(a,3)
@

@@

We use SAE to estimate the subpopulation ratios, and then use the \code{simputation}
package \citep{loo2022simputation} to impute the missing values. 
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = min_complete(n = 10, vars = "industrial")
         , R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
head(r,3)
@
To impute the values, using \code{impute_proxy()} we need to merge the ratios
with the producers dataset (which automatically happens by SBI and size class).
<<>>=
library("simputation")
dat <- merge(producers, r)
dat <- impute_proxy(dat, industrial ~ R*total)
@
We can inspect the imputed values as follows.
<<>>=
iNA <- is.na(producers$industrial)
head(dat[iNA, c("sbi", "size", "level", "R", "industrial", "total")])
@
From this, we get all the information to interpret the imputed values. For
example, we see that in record 664, the ratio was estimated after collapsing
the group $(\texttt{sbi},\texttt{size})=(21122,7)$ to $\texttt{sbi2}=21$ since
the collapse \code{level} equals 2.

@@

We first define an `aggregation' function that takes a vector 
of donors and returns a non-empty sample.
<<>>=
random_element <- function(x) sample(x[ !is.na(x)], 1)
@
We will use \code{cumulate()} to ensure that there are at least five non-empty
values in \code{x} when \code{random_element()} is called. To make sure we
obtain a donor for each record, we add an identifying column \code{id} to use
as grouping variable.  
<<>>=
producers <- cbind(id = sprintf("ID%03d", seq_len(nrow(producers)))
                  , producers)
set.seed(111)
imputations <- cumulate(producers
                 , collapse = id ~ sbi*size + sbi + sbi2 + sbi1
                 , test = min_complete(5,"trade")
                 , donor_trade = random_element(trade))

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing Complex Aggregates} \label{sect:complex}
Until now, the aggregates have been simple (scalar) values.  With the
\code{cumulate()} function it is also possible to specify complex aggregates that
go beyond simple aggregates. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = min_complete(n = 10, vars = c("total","industrial"))
         , model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.

>>>>

cumulate(input, collapse = A * B ~ A * B1 + A
        , test = function(d) nrow(d) >= 3, muY = mean(Y) )
@
%
Consider the formula \code{A * B ~ A * B1 + A} in the call to \code{cumulate()}.
The left-hand-side \code{A * B} is the target output grouping. The
right-hand-side is to be interpreted as the collapsing sequence: if an instance
of \code{A * B} does not pass the test, then collapse to \code{A * B1}, and if that
does not pass the test collapse to \code{A}. If this final grouping also does
not pass the test, the result is \code{NA}.

@@

It is possible to get the same result with \code{accumulate()}. This will cause
summation over all variables that are not used in the formula object. In the
below example we also introduce the helper function \code{min_records()}.
<<>>=
input$Y2 <- 11:19
accumulate(input, collapse = A * B ~ A * B1 + A
          , test = min_records(3), fun = mean)
@
This means that for experimentation, users must be careful to exclude categorical
variables that are not used from the input data set. For example, if \code{B1}
is not used, we get the following.
<<>>=
accumulate(input[-3], collapse = A * B ~ A
          , test = min_records(3), fun = mean)
@

@@

We will apply these conditions to the \code{input} dataset that was 
constructed in Section~\ref{sect:formula}. As a reminder we print
the first few records.
<<>>=
head(input, 4)
@
We use \code{A * B ~ A * B1 + B1} as collapsing scheme. The function
\code{from_validator} passes the requirements as a test function to
\code{accumulate()} (or \code{cumulate()}).
<<>>=
accumulate(input, collapse = A * B ~ A * B1 + B1,
    test = from_validator(rules), fun = mean)
@
Note that for target groups $(A=3,B=21)$ and $(A=3,B=22)$ none of the available
collapsing levels lead to a group that satisfied all conditions. Therefore the
collapsing level and output variables are all missing (\code{NA}).

The third and most flexible way for users to express tests is to write a custom
testing function. The requirements are that it must work on any subset of a
data frame, including a dataset with zero rows. The previous example can
thus also be expressed as follows.
<<>>=
my_test <- function(d) nrow(d) >= 3 && sum(d$Y >= 2) >= 3
accumulate(input, collapse = A * B ~ A * B1 + B1, test = my_test, fun = mean)
@

@@

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation that includes the target subpopulation
\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
by simple random sampling from the population, the mean can be estimated with
the sample mean. In this example we will demand that there are at least ten
records for which turnover has been measured. The collapsing scheme is given by
\code{sbi * size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
classification by respectively the first two SBI digits and the first SBI
digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi, 1, 2)
producers$sbi1 <- substr(producers$sbi, 1, 1)
head(producers, 3)
@
Using \code{cumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers
             , collapse = sbi * size ~ sbi + sbi2 + sbi1
             , test = min_complete(n = 10, vars = "industrial")
             , mean_industrial = mean(industrial, na.rm = TRUE))
head(a,3)
@

@@

We use SAE to estimate the subpopulation ratios, and then use the \code{simputation}
package \citep{loo2022simputation} to impute the missing values. 
<<>>=
r <- cumulate(producers
         , collapse = sbi * size ~ sbi + sbi2 + sbi1
         , test = min_complete(n = 10, vars = "industrial")
         , R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
head(r,3)
@
To impute the values, using \code{impute_proxy()} we need to merge the ratios
with the producers dataset (which automatically happens by SBI and size class).
<<>>=
library("simputation")
dat <- merge(producers, r)
dat <- impute_proxy(dat, industrial ~ R * total)
@
We can inspect the imputed values as follows.
<<>>=
iNA <- is.na(producers$industrial)
head(dat[iNA, c("sbi", "size", "level", "R", "industrial", "total")])
@
From this, we get all the information to interpret the imputed values. For
example, we see that in record 664, the ratio was estimated after collapsing
the group $(\texttt{sbi},\texttt{size})=(21122,7)$ to $\texttt{sbi2}=21$ since
the collapse \code{level} equals 2.

@@

We first define an `aggregation' function that takes a vector 
of donors and returns a non-empty sample.
<<>>=
random_element <- function(x) sample(x[ !is.na(x)], 1)
@
We will use \code{cumulate()} to ensure that there are at least five non-empty
values in \code{x} when \code{random_element()} is called. To make sure we
obtain a donor for each record, we add an identifying column \code{id} to use
as grouping variable.  
<<>>=
producers <- cbind(id = sprintf("ID%03d", seq_len(nrow(producers)))
                  , producers)
set.seed(111)
imputations <- cumulate(producers
                 , collapse = id ~ sbi * size + sbi + sbi2 + sbi1
                 , test = min_complete(5,"trade")
                 , donor_trade = random_element(trade))

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing Complex Aggregates} \label{sect:complex}
Until now, the aggregates have been simple (scalar) values.  With the
\code{cumulate()} function it is also possible to specify complex aggregates that
go beyond simple aggregates. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi * size ~ sbi + sbi2 + sbi1
         , test = min_complete(n = 10, vars = c("total","industrial"))
         , model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.
